Importance of using the correct background list for over-representation analysis

### 2024-05-31 old/analysis/example_proteomics.Rmd
move

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("prot_kegg_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("prot_reactome_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

>>>>

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Robust detection according to the study authors |
| Gene Selection Criteria | limma FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

ora_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Robust detection according to the study authors |
| Gene Selection Criteria | limma FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("prot_kegg_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Robust detection according to the study authors |
| Gene Selection Criteria | limma FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Robust detection according to the study authors |
| Gene Selection Criteria | limma FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("prot_reactome_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

### 2024-05-31 old/analysis/example_rnaseq.Rmd
move

png("rna_kegg_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

png("rna_reactome_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

>>>>

svg("rna_kegg_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

svg("rna_reactome_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

### 2024-05-31 old/analysis/example_scrnaseq.Rmd
move

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | Muscat FDR<0.05 CD4 T-cells |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 5 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("scrna_kegg_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("scrna_reactome_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

>>>>

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Muscat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 5 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Muscat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("scrna_kegg_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Muscat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Muscat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("scrna_reactome_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

### 2024-05-31 old/manuscript/manuscript.Rmd
move

>>>>

## Methods

### Demonstrating the impact of the problem

To demonstrate the impact of this problem on real omics data, we conducted enrichment analysis of
four different types of data.
We obtained differential proteomics data from Stukalov et al (https://pubmed.ncbi.nlm.nih.gov/33845483/),
who profiled the effect of SARS-COV2 infection on A549 cells.
At 24 hrs infection with SARS-CoV2, Stukalov et al reported the detection of 5825 proteins, including
165 downregulated and 44 upregulated (FDR<0.05).
For bulk RNA-seq, we obtained count data from DEE2.io (Ziemann et al, 2019) for control and azacitidine
exposed AML3 cells (SRA accession SRP038101) (Lund et al, 2014, PMID:25315154).
Genes with fewer than 10 reads per sample on average were considered not detected and were omitted from
downstream analysis.
DESeq2 was used to compare treatment groups and genes with FDR<0.05 were selected (Love 2014).
For single cell RNA-seq 10X Genomics droplet-based, scRNA-seq PBMC data from 8 Lupus patients obtained
before and after 6h-treatment with IFN-β (Kang et al. 2018).
The complete raw data, as well as gene and cell metadata is available through the NCBI GEO,
accession number GSE96583.
Analysis of this data set  was done using a pseudobulk approach according to the Muscat package vignette
(Crowell et al, 2020), yielding lists of detected genes for each cell state as well as up- and
down-regulated gene lists (FDR<0.05).
Infinium 450K BeadChip based Epigenenome Wise Association Study (EWAS) data was obtained from
Joehanes et al (PMID:27651444), a meta-analysis investigating the DNA methylation differences between
smokers and non-smokers.
Data was downloaded from EWAS Catalog at the link (http://ewascatalog.org/?study=27651444_smoking_current_vs_never_smoking).
As a background gene list, all genes represented by probes on the 450K array were listed.
For all of the above analyses, up- and down-regulated genes were partitioned into separate lists,
for separate enrichment analysis.
These gene lists were loaded into R and were subject to ClusterProfiler analysis along with the
respective background gene set.
KEGG (REF) or Reactome (REF) pathway annotations were used (GMT format).
KEGG pathway annotations v2023.1 were obtained from MSigDB (REF), while Reactome pathways were downloaded
from the Reactome website (2023-03-06).
Pathways (gene sets) with fewer than 10 detected genes were not included in the analysis.
In addition to the default ClusterProfiler analysis, a modification was made to circumvent the
background filtering problem.
To do this, the list of detected genes was appended as a new "pathway" to the KEGG or Reactome annotation
set.
The results of this "fixed" analysis was compared those using the default algorithm using an Euler
diagram.

### Examining the breadth of the problem

Next, we investigated the behaviour of many different ORA tools to find out whether they suffered from
the same problem or not.
We analysed the 20 most popular ORA tools as described by Xie et al, 2021 (PMID:33858350).
To understand whether the background list filtering problem was apparent, we used the gene lists from
the bulk RNA-seq dataset, with various pathway /gene set databases.
If the tool reported the number of genes in the background list, we used this as evidence to classify
the tool's behaviour.
If not, then we undertook forensic investigation of the enrichment results to infer the program's
behaviour.

## Results

### Background list filtering biases ORA results

### Inappropriate background list filtering is widespread
